Tree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest

### 2019-10-20 manuscript/climate_sensitivity_SI.Rmd

heights <- read.csv("tables_figures/height_regression.csv", stringsAsFactors = FALSE)
kable(heights, booktabs=TRUE, format = "latex", caption = "Species-specific height regression equations") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

>>>>

heights <- read.csv("tables_figures/height_regression.csv", stringsAsFactors = FALSE)
heights[,2] <- NULL
kable(heights, booktabs=TRUE, format = "latex", caption = "Species-specific height regression equations") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

### 2019-10-20 manuscript/climate_sensitivity_draft.Rmd

(**add description of climate data used in Fig. 1, NEON vertical profiles**)

@@

We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events where tree growth was substantially reduced and where peak growing season climatic conditions were among the driest of the time period. To identify years with widespread reductions in tree growth, we used the pointRes package [@R-pointRes] in R (version 3.5.3) to determine drought periods based on trees' drought resistance, which is defined as the ratio between the performance during and before the disturbance [@lloret_components_2011]. Specifically, we looked at the ratio between annual basal area increment (BAI) in the year of the drought to average annual BAI in the 5 preceding years. Candidate drought years were defined if >25% of the cored trees experienced <30% growth in a year compared to the previous 5 years. Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the months (of the current year) to which annual growth was most sensitive for trees at this site. We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between potential evapotranspiration (PET) and precipitation (PRE). These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris etal., 2014). The driest years were identified through simply ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.

@@

The droughts considered here were of similar severity (Fig. 1b) and fairly moderate; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019). Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature. While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), underlining the fact that these droughts did not induce extreme stress on the entire forest. It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 10-12% for all droughts combined, and 21-27% for each individual drought (Table 5). Methodologically, the moderate nature of these drougths is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed. These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*). Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size (Bennett et al. 2015; Stovall et al. 2019) and--to some extent-- traits (**REFS**). Nevertheless, our results are consistent with findings from more extreme droughts.

Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height. To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (**REFS**). They also face different microenvironments, which are partially related to canopy position. Even under non-drought conditions, evaporative demand increases with tree height, which are more closely coupled to the atmosphere  (Fig. 1a-b; **REFS- Jarvis 1984?**; [@bretfeld_plant_2018]). **The vegetation reaches higher temperatures[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)**, which would occur more during a drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. Furthermore, daytime CO_2 concentrations tend to be lower in the canopy than in the understory **(Fig. 2[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)** ([Koike et al. 2001](https://watermark.silverchair.com/21-12-13-951.pdf); MORE REFS), implying greater water costs of $CO_2$ uptake. **[McDowell-vertical profiles in isotopes?*]**. Taller trees are more likely to be in domiant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig. 2**e?**) [@muller-landau_testing_2006]. We show that height is a far stronger predictor than crown position (Tables 1,4,5). Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4). However, rates of change are relatively slow, and the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of (*H1.2*), implying that dominant crown positions did indeed reduce $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??).  It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer distproportionately during drought (**REFS**). The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.

Beyond microenvironment, multiple tree characteristics covary with height. To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including crown volume, rooting volume, sapwood area, transpiration ... (e.g., [Anderson-Teixeira et al. 2015_scaling](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.12470); **REFS--KAT: Sperry scaling model; roots reference; BAAD database?**). These work in combination to shape whole-plant hydraulic function (**REFS-KAT: Sperry scaling model, check McDowell scaling paper, Couvreur et al. 2018; others?**), and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees. However, height is arguably the most biophysically meaningful tree-size variable shaping drought response. Beyond whole-plant characteristics, leaf traits vary with height on the tree--sometimes quite dramatically, with upper-canopy leaves ... (**describe differences**) ([Couvreur et al. 2018](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/pce.13322?casa_token=92wZmmL9EhQAAAAA%3A2cq5yAR0_pyM9TVBa5o_DKkhDuHwIUFryxZ43zsdTgQyEa8f0ZfiEHZ5SHV7fFB49UkzXNvL7nMwabU); [Koike et al. 2001](https://watermark.silverchair.com/21-12-13-951.pdf)). Our analysis focuses on species-level comparisons and does not characterize the role of variation with height. On an individual level, average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits--assuming trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing drought resistance with height. In not accounting for this, our analysis is most likely biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$. Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses (*Couvreur et al. 2018*).

@@

Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydaulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importantance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5). [**BRIEFLY DISCUSS DIFFERENCES ACROSS DROUGHTS-- For example, 1966 was preceded by two dry years, and height had the strongest effect in this drought, consistent with the finding that height becomes a stronger predictor of mortality as drought duration and intensity increase (Stovall et al. 2019). ...**] Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites would be of great value to elucidating the mechanisms through which these shape drought resistance (**check for refs**). Unfortunately, our ability to compare forest responses across droughts is hampered by the fact that droughts are rarely explicitly defined in ecological studies [@slette_how_2019]. Further research will be required to understand how drought characteristics interact with driver variables to shape tree growth responses.

>>>>

To determine a vertical height climate profile, data was measured by a National Ecological Observation Network (NEON) eddy-covariance (?) flux tower located within a 1km from the study area. In total, wind, relative humidity, and air temperature data were collected for the years 2016-2018 [@noauthor_national_nodate]. After filtering for missing and outlier values, the data was consolidated to represent the mean values per sensor height per day. The resultant plots (Figure 2) illustrate the maximum and minimum of these mean values, with error represented by the standard deviation.

@@

We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events where tree growth was substantially reduced and where peak growing season climatic conditions were among the driest of the time period. To identify years with widespread reductions in tree growth, we used the pointRes package [@R-pointRes] in R (version 3.5.3) to determine drought periods based on trees' drought resistance, which is defined as the ratio between the performance during and before the disturbance [@lloret_components_2011]. Specifically, we looked at the ratio between annual basal area increment (BAI) in the year of the drought to average annual BAI in the 5 preceding years. Candidate drought years were defined if >25% of the cored trees experienced <30% growth in a year compared to the previous 5 years. Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the months (of the current year) to which annual growth was most sensitive for trees at this site. We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between potential evapotranspiration (PET) and precipitation (PRE). These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014). The driest years were identified through simply ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.

@@

The droughts considered here were of similar severity (Fig. 1b) and fairly moderate; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019). Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature. While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), underlining the fact that these droughts did not induce extreme stress on the entire forest. It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 10-12% for all droughts combined, and 21-27% for each individual drought (Table 5). Methodologically, the moderate nature of these drougths is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed. These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*). Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent-- traits (**REFS**). Nevertheless, our results are consistent with findings from more extreme droughts.

Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height. To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (**REFS**). They also face different microenvironments, which are partially related to canopy position. Even under non-drought conditions, evaporative demand increases with tree height, which are more closely coupled to the atmosphere  (Fig. 1a-b; **REFS- Jarvis 1984?**; [@bretfeld_plant_2018]). **The vegetation reaches higher temperatures[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)**, which would occur more during a drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. Furthermore, daytime CO_2 concentrations tend to be lower in the canopy than in the understory **(Fig. 2[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)** [@koike_leaf_2001; MORE REFS], implying greater water costs of $CO_2$ uptake. **[McDowell-vertical profiles in isotopes?*]**. Taller trees are more likely to be in domiant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig. 2**e?**) [@muller-landau_testing_2006]. We show that height is a far stronger predictor than crown position (Tables 1,4,5). Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4). However, rates of change are relatively slow, and the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of (*H1.2*), implying that dominant crown positions did indeed reduce $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??).  It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer distproportionately during drought (**REFS**). The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.

Beyond microenvironment, multiple tree characteristics covary with height. To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including crown volume, rooting volume, sapwood area, transpiration ... (e.g., [@andersonteixeira_size-related_2015; **REFS--KAT: Sperry scaling model; roots reference; BAAD database?**)]. These work in combination to shape whole-plant hydraulic function (**REFS-KAT: Sperry scaling model, check McDowell scaling paper, Couvreur et al. 2018; others?**), and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees. However, height is arguably the most biophysically meaningful tree-size variable shaping drought response. Beyond whole-plant characteristics, leaf traits vary with height on the tree--sometimes quite dramatically, with upper-canopy leaves ... (**describe differences**) [@couvreur_water_2018; @koike_leaf_2001]. Our analysis focuses on species-level comparisons and does not characterize the role of variation with height. On an individual level, average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits--assuming trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing drought resistance with height. In not accounting for this, our analysis is most likely biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$. Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses [@couvreur_water_2018].

@@

Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydaulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importantance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5). [**BRIEFLY DISCUSS DIFFERENCES ACROSS DROUGHTS-- For example, 1966 was preceded by two dry years, and height had the strongest effect in this drought, consistent with the finding that height becomes a stronger predictor of mortality as drought duration and intensity increase** [@stovall_tree_2019]. ...] Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites would be of great value to elucidating the mechanisms through which these shape drought resistance (**check for refs**). Unfortunately, our ability to compare forest responses across droughts is hampered by the fact that droughts are rarely explicitly defined in ecological studies [@slette_how_2019]. Further research will be required to understand how drought characteristics interact with driver variables to shape tree growth responses.
